plyranges: a grammar for transforming genomics data

### 2018-05-08 paper.Rmd

Since the GRanges object is a tabular data structure our grammar includes
operators to filter, sort and aggregate by columns in a GRanges. These 
operations can be performed over partitions within GRanges object
using the *group_by* modifier. Together with our algebra for arithmetic and
merging, these operations enable flexible and reproducible analysis using
the desgin principles of the `dplyr` grammar.

## Cohesion

A function is cohesive if it performs a singular task and does not produce
any side-effects. In the `plyranges` DSL our algebra for performing
genomic arithmetic is a key example of cohesion. We define anchoring operators
that decorate a GRanges object with an 'anchor' and modify the semantics
of performing genomic arithmetic. The anchoring operators amount to fixing a GRanges
object by its start, center, or end coordinates (or fixing these coordinates
by strand). This enables any arithmetic function that performs a 
coordinate transformation to remain cohesive: that is they always perform
the same operation regardless of whether a GRanges object is anchored or not.
The only difference is the result of performing the arithmetic changes when
contextual information is added to the GRanges object.

Another example of our algebra altering object semantics, while maintaining 
cohesion is through the 'group_by' operator. Like anchoring, this operator 
decorates a GRanges object with a column name (or names) that defines a 
partitioning of the  GRanges by the unique values in the column(s). Functions 
defined in the  `plyranges` DSL that perform restriction, aggregation, or 
column modification still perform those singular tasks, however grouping changes
how those tasks are performed.

## Consistency

A core design principle of the `plyranges` DSL is interface consistency:
a user should not be surprised by the input or output of `plyranges` code.
A key example of consistency is how `plyranges` handles strand information.
Every function that computes with strand information indicates its intentions
by including suffixes such as 'directed', 'upstream' or 'downstream' in its
name, otherwise strand is ignored. This strongly differs from 
the Bioconductor packages, which produces surprising output by assuming the 
user is always interested in using strand in the majority of circumstances 
(unless a user is computing coverage or finding flanking ranges).

Our use of suffixes in function names, also highlights a core tenet of 
the `plyranges` design: avoid complex generic functions with many arguments
and instead use cohesive functions with a minimal number of arguments. As an
example, we have written a consistent framework for reading and writing 
files from and to common genomic data formats as GRanges, 
using the `rtracklayer` package as a back-end @Lawrence2009-nt. We have replaced
that packages generic functions for importing files with a family of reader
functions that all take the exact same arguments.

## Endomorphism

Most function calls in `plyranges` are endomorphisms: when the input 
is GRanges object the output will also be a GRanges object. The use of
endomorphism in the `plyranges` DSL enables a user to predict the structure of 
the output of their computations and does not require them to learn any 
additional classes beyond GRanges and DataFrames.

As an example, in `plyranges` when we compute coverage over a GRanges object,
the result of the operation is an expanded GRanges object with a score column,
corresponding the estimated coverage over a genomic region. Hence, the resulting
coverage score can easily be composed with other expressions in our algebra.
This pattern strongly deviates from the design of the OO interface in 
`GenomicRanges`, which in the case of computing coverage would return an RleList,
an object that has many new methods and that is potentially unfamiliar to 
a new user. While, these low-level classes enable efficient computing,
the increase complexity, and as a consequence are abstracted away in the 
`plyranges` DSL.

## Fluency

As a consequence of the design principles defined above every function in `plyranges`
performs a single action on GRanges objects. The `plyranges` DSL implements the 
core verbs from the `dplyr` package and implements a genomic relational algebra
for transforming GRanges objects (table x, y). Each verb preserves the 
semantics of GRanges object and works with derivatives of the GRanges class. 
Both of these aspects reduce the cognitive load on a new user since most 
manipulations can be performed with a vocabulary of several verbs, rather 
than having to memorise function names that are nouns.

This approach strongly contrasts the `GenomicRanges/IRanges` OO interface, 
which emphasises the use of setter and getter methods. In that interface, 
core components and metadata are updated via replacement methods 
(requiring knowledge of the class components), while our interface requires 
only a single call to `mutate()` to perform the modification.

Workflows can be composed by chaining verbs together into 'sentences' via the 
forward pipe operator,`%>%`  (exported from the R package `magrittr` 
@R-magrittr), which can be read as the word 'then'. Overall, this allows users 
to write human readable code because workflows describe what the code is doing 
rather than how its doing it.

## Opportunities

@@

>>>>

Since the GRanges object is a tabular data structure, our grammar includes
operators to filter, sort and aggregate by columns in a GRanges. These 
operations can be performed over partitions formed
using the *group_by* modifier. Together with our algebra for arithmetic and
merging, these operations conform to the semantics and syntax of the 
`dplyr` grammar.

## Design principles

The design of plyranges adheres to well understood principles of
language and API design: cognitive consistency, cohesion,
expressiveness and endomorphism [cite: Green's cognitive
dimensions]. To varying degrees, these principles also underly the
design of dplyr and the Bioconductor infrastructure.

### Cognitive consistency and fluency

We aim for our interfaces to have a simple and direct mapping to the
user's cognitive model, i.e., how the user thinks about the data. This
requires careful selection of the level of abstraction so that the
user can express workflows in the language of genomics. This motivates
the adoption of the tidy GRanges object as our central data
structure. The basic data.frame and dplyr tibble lack any notion of
genomic ranges and so could not easily support our genomic grammar,
with its specific verbs for range-oriented data manipulation. Another
example of cognitive consistency is how plyranges is insensitive to
direction/strand by default when, e.g., detecting
overlaps. GenomicRanges has the opposite behavior. We believe that
defaulting to purely spatial overlap is most intuitive to most users.

<!--
ML:
Hadley actually presents dplyr as a 'cognitive tool', i.e., something
that helps users think about the problem. In my opinion, the best we
can do is attempt to match the intuition of the user, so when the user
decides what to do, there is an obvious way of doing it. I'm not ready
to claim that the user should start from our grammar.
-->

Like dplyr, plyranges verbs are functional: they are free of side
effects and return their result. This enables chaining of verbs
through syntax like the forward pipe operator (`%>%`, read as "then")
of the magrittr package @R-magrittr. This has syntax has a direct
cognitive mapping to natural language and the intuitive notion of
pipelines. The low-level object-oriented APIs of Bioconductor tend to
manipulate data via subreplacement functions, like `start(gr) <-
x`. These ultimately produce the side effect of replacing a symbol
mapping in the current environment and thus are not amenable to
so-called fluent syntax.

### Cohesion

A function is cohesive if it performs a singular task without
producing any side-effects. Singular tasks are not necessarily atomic;
they can always be broken down further at lower levels of abstraction.
For example, to resize a range, the user needs to specify which
position (start, end, midpoint) should be invariant over the
transformation. The `resize()` function from the GenomicRanges package
has a `fix` argument that sets the anchor, so calling `resize()`
coalesces anchoring and width modification. The coupling at the
function call level is justified since the effect of setting the width
depends on the anchor. However, plyranges increases cohesion and
decouples the anchoring into its own function call.

Increasing cohesion simplifies the interface to each operation, makes
the meaning of arguments more intuitive, and relies on function names
as the primary means of expression, instead of a more complex mixture
of function and argument names. Since functions are superordinate to
their arguments, flattening the API at the function level enables the
user to conceptualize the API as a catalog of functions, without
having to descend further. A flat function catalog also enhances API
discoverability, particularly through auto-completion in IDEs. One
downside of pushing cohesion to this extreme is that function calls
become coupled, and care is necessary to treat them as a group when
modifying code.

### Expressiveness

Expressiveness relates to the information content in code: the
programmer should be able to clarify intent without unnecessary
verbosity. For example, our overlap-based join operations are more
concise than the multiple steps necessary to achieve the same effect
in the original GenomicRanges API. In other cases, the plyranges API
increases verbosity for the sake of clarity and cohesion. Explicitly
calling `anchor()` can require more typing, but the code is easier to
comprehend. Another example is the set of routines for importing
genomic annotations, including `read\_gff()`, `read\_bed()`, and
`read_bam()`. Compared to the generic `import()` in rtracklayer, the
explicit format-based naming in plyranges clarifies intent and the
type of data being returned. Similarly, every plyranges function that
computes with strand information indicates its intentions by including
suffixes such as `directed`, `upstream` or `downstream` in its name,
otherwise strand is ignored. The GenomicRanges API does not make this
distinction explicit in its function naming, instead relying on a
parameter that defaults to strand sensitivity, an arguably unintuitive
behavior.

@@

# Discussion
